---
title: "project"
author: "sai"
date: "10/01/2022"
output: html_document
---

# Divvy_Exercise_Full_Year_Analysis

## This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"
# Installing required packages
## tidyverse for data import and wrangling
## lubridate for date functions
## ggplot for visualization
```{r}
install.packages("tidyverse",repos = "http://cran.us.r-project.org")

library(tidyverse)
library(lubridate)
library(ggplot2)
```

#setting path
```{r}
getwd()
setwd("C:/Users/Erumalla Saikumar/OneDrive/Documents/google capstone")

```


# Uploading Divvy datasets (csv files)
```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```



# Comparing column names each of the files
##While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file
```{r}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)
```


# Renaming the columns to make them consistent
```{r}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))
```


# Inspect the dataframes and look for incongruencies
```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```



# Converting ride_id and rideable_type to character so that they can stack correctly
```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
q1_2020 <-  mutate(q1_2020,  started_at = as.POSIXct(started_at, format = "%Y-%m-%dT%H:%M:%S+000"))
q1_2020 <-  mutate(q1_2020,  ended_at = as.POSIXct(ended_at, format = "%Y-%m-%dT%H:%M:%S+000"))
```

                                     


# Stack individual quarter's data frames into one big data frame
```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```



# emoving extra columns
## Removing lat, long, birthyear, and gender fields as this data was dropped beginning in 2020
```{r}
colnames(all_trips)
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```




# Inspecting the new table that has been created

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)
```

# There are a few problems we will need to fix:
## (1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
## (2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
## (3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
## (4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides
## In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
## Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature
```{r}
table(all_trips$member_casual)
```


# Reassigning to the desired values (we will go with the current 2020 labels)
```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```


```{r}
table(all_trips$member_casual)
```

#Adding columns
## Adding columns that list the date, month, day, and year of each ride
## This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level
```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```


# Adding a "ride_length" calculation to all_trips
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```



# Inspecting the structure of the columns
```{r}
str(all_trips)
```


# Converting "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```




# Removing "bad" data
## The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
## We will create a new version of the dataframe (v2) since data is being removed
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

all_trips_v2<- na.omit(all_trips_v2)
```


# Descriptive analysis on ride_length
```{r}
summary(all_trips_v2$ride_length)
```



# Comparing members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```




# Seeing the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```




# Notice that the days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


# Now, let's run the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```


#visualising data

## analyzing ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)
```

	


# Let's visualize the number of rides by rider type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```



# Let's create a visualization for average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```





#Exporting the data
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'C:/Users/Erumalla Saikumar/OneDrive/Documents/google capstone/avg_ride_length.csv')
```

#MY FINDINGS
##In conclusion, based on the above analysis, there are no clear recommendations on how to convert casual cyclists to members. But, there are some follow-up leads which will want some more further investigation which may lead to conversion.

##Firstly as there is no personal information available there are several limitations in the data provided. We are unable to tell how often the same user uses the bike in the month and how frequently, and from which stations. Next, there are blank data points for start & end stations and starting and ending longitudes & latitudes. This is a data issue which requires further investigation as there are quite a high number of such missing data points. With more complete data, we can have a better analysis.

##From the data analysis, we can observe that there are a few start and end recurring stations with higher footfall among the casual cyclists. We can put up targeted advertisements on those stations, or have a personalized advertisement via the cycling app for discounted membership subscription promotions for cyclists using those stations more frequently..

##There is also a high number of casual cyclists who take abnormally long trips. This requires further investigation to understand why is this the case, and hopefully draw some insights which may help to convert them to members.

##Lastly, there is an increased bike usage during the weekends (Friday – Sunday) among the casuals. We might increase rental cost over the weekends for the single and multi passes such that it might be more worthwhile for the cyclists to switch to member subscription instead.













